---
title: "Pedigree-based BLUP"
author: "Gota Morota"
date: "March 6, 2018"
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: false
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Overview  
We will learn how to connect phenotypes and pedigree using pedigree-based best linear unbiased prediction (PBLUP). 

# Reading R object
Use the function `read_excel` in the `readxl` package to read the pedigree file (Simdata.xlsx) in a data frame format. Use the function `load()` to load the numerator relationship matrix (`A.Rda`) we created in class ([link1](http://morotalab.org/asci431-2018/day06/day06.html)).
```{r, eval=FALSE}
library(readxl)
dat <- read_excel(file.choose())
dim(dat)
head(dat)

load(file.choose()) # A.Rda
dim(A)
```

# Statistical model 
The statistical model we fit is given by 
$$
\mathbf{y} = \mathbf{X}\mathbf{b} + \mathbf{Z}\mathbf{u} + \mathbf{e},
$$
where $\mathbf{y}$ is the vector of birth weights,  $\mathbf{X}$ and $\mathbf{Z}$ are incident matrices for fixed and random effects, respectively, $\mathbf{b}$ is the vector of regression coefficients for fixed effects, $\mathbf{u}$ is the vector of regression coefficients for random genetic values, and $\mathbf{e}$ is the vector of residuals. 
Recall that BLUP of $\mathbf{u}$ is given by Henderson (1975) ([doi: 10.2307/2529430](https://www.jstor.org/stable/2529430))
$$
\begin{align*}
BLUP(\mathbf{u}) &= E(\mathbf{u} | \mathbf{y}) \\
&= Cov(\mathbf{u}, \mathbf{y}') Var(\mathbf{y})^{-1} [\mathbf{y} - E(\mathbf{y})]   \\
&= \mathbf{A}\sigma^2_{A} \mathbf{Z}'\mathbf{V}^{-1}(\mathbf{y} - \mathbf{X}\hat{\mathbf{b}}),
\end{align*}
$$
where 
$$
\begin{align*}
\mathbf{V} &= Var(\mathbf{y}) \\
&= \mathbf{Z}\mathbf{A}\sigma^2_A\mathbf{Z}'  + \mathbf{I}\sigma^2_e
\end{align*}
$$

We predict estimated breeding values (EBV) or genetic values $\hat{\mathbf{u}}$ in the following two steps. 

* fit OLS to estimate fixed effects ($\hat{\mathbf{b}}$)
* fit BLUP to obtain EBV ($\hat{\mathbf{u}}$) condition on the estimated $\hat{\mathbf{b}}$

Later we will discuss how to obtain $\hat{\mathbf{b}}$ and $\hat{\mathbf{u}}$ simultaneously. 

# Incidence matrix X
We will now contruct each component one by one. First we will create the incidence matrix $\mathbf{X}$. 
We first subset the variable `dat` by age of dams and sex of calves, and create a new variable `dat2`. 
The `model.matrix()` function returns the incidence matrix $X$. 
```{r, eval=FALSE}
dat2 <- dat[, c("AgeDam", "Sex")]
head(dat2)
X <- model.matrix(~ dat2$AgeDam + dat2$Sex)
dim(X)
head(X)
tail(X)
```

# Incidence matrix Z
Next we will create the incidence matrix $\mathbf{Z}$. 
```{r, eval=FALSE}
Z <- diag(nrow(A))
dim(Z)
diag(Z)
```

# Incidence matrix I
The third incidence matrix we create is $\mathbf{I}$. 
```{r, eval=FALSE}
I <- diag(nrow(A))
dim(I)
diag(I)
```

# Variance components estimation using the regress package
The `regress()` function in the [regress](https://cran.r-project.org/web/packages/regress/index.html) package fits a liner mixed model and estimate variance components ($\sigma^2_A$ and $\sigma^2_e$) through a restricted maximum likelihood (REML). The variance-covariance structure of $\mathbf{u}$ is given by the numerator relationship matrix $\mathbf{A}$. 
```{r, eval=FALSE}
install.packages("regress")
library(regress)
`?`(regress)

y <- dat$BW_205d
varcomp <- regress(y ~ -1 + X, ~A)
summary(varcomp)
varcomp$sigma

sigma2A <- varcomp$sigma[1] # additive genetic variance
sigma2A
sigma2e <- varcomp$sigma[2] # residual variance 
sigma2e
```

### Exercise 1
What is the estimate of heritability? Compare the estimate with the one we obtained from [intraclass correlation](http://morotalab.org/asci431-2018/day13/day13.html) using the sire model. 
```{r, eval=FALSE, echo=FALSE}
h2A <- sigma2A / (sigma2A + sigma2e)
h2A
```

# Inverse of V
Here we compute the inverse of $V$ matrix. We will use 1) the multiplication operator `%*%` and 2) the function `solve()` to obtain the inverse of matrix. 
```{r, eval=FALSE}
V <- Z %*% A %*% t(Z) * sigma2A + I * sigma2e
dim(V)
Vinv <- solve(V) 
dim(Vinv)
```

# Ordinary least squares
We will use the `lm()` function to fit OLS and estimate fixed effect coefficients. This function offers an intuitive formula syntax. The `summary()` function summarizes a model fit. 
```{r, eval=FALSE}
fit <- lm(BW_205d ~ AgeDam + factor(Sex), data=dat)
summary(fit)
residuals(fit)
```

# PBLUP of genetic values
Now we compute EBV ($\hat{\mathbf{u}}$). 
```{r, eval=FALSE}
uhatA <- sigma2A * A %*%  t(Z) %*% Vinv %*%  (matrix(y) - matrix(predict(fit)))
uhatA2 <- sigma2A * A %*%  t(Z) %*% Vinv %*%  (matrix(y) - matrix(X%*% fit$coefficients)) # alternative
table(uhatA == uhatA2)
head(uhatA)
tail(uhatA)
```

Let's plot a histogram of EBV. 
```{r, eval=FALSE}
hist(uhatA, col = "lightblue", main="Histogram", xlab="Estimated breeding values")
```

### Exercise 2
Create a scatter plot of EBV vs. observed values. Interpret the result. 
```{r, eval=FALSE, echo=FALSE}
plot(y, uhatA, xlab="Estimated breeding values", ylab="Observed values")
abline(lm(uhatA ~ y), lwd=3, col="red") 
```

Note that the estimate of heritability can also be obtained by regressing `uhatA` on `y`.
```{r, eval=FALSE, echo=TRUE}
summary(lm(uhatA ~ y))
```

### Exercise 3
Which individual has the highest EBV? Which individual has the lowest EBV? Use the `which.max()` and `which.min()` functions.
```{r, eval=FALSE, echo=FALSE}
which.max(uhatA)
which.min(uhatA)
```


### Exercise 4
Rank animals based on their EBV. Show the top six animal IDs and their EBV. Use the `order()` function.
```{r, eval=FALSE, echo=FALSE}
uhatA.rank <- uhatA[order(uhatA, decreasing = TRUE)]
head(uhatA.rank)
uhatA.rank.names <- dat$Animal[order(uhatA, decreasing = TRUE)]
head(uhatA.rank.names)
```

### Exercise 5
If you are going to select only top 10% of animals, which animals would you pick? 
```{r, eval=FALSE, echo=FALSE}
1004 * 0.1
uhatA.rank[1:100]
uhatA.rank.names[1:100]

hist(uhatA, col = "lightblue", main="Histogram", xlab="Estimated breeding values")
abline(v=uhatA.rank[100], lwd=3, col="red")
```


# Save as R object
```{r, eval=FALSE}
save(uhatA, h2A, sigma2A, file="day17.Rda")
```



